Hijacking Simulators with Universal Probabilistic Programming

### 2019-05-19 paper/neurlps2019/main.tex
updated

Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way. Thus, simulators by design are ideal programs for probabilistic programming systems~(PPSs).
However, within existing PPSs one would have to re-implement the simulator via the PPS language specification, which is inefficient and often not feasible due to the complexity of such scientific and industrial simulators.
However, recent work
by \cite{baydin2018efficient} demonstrated a pathway to turn a particular
type of event-based simulator into a probabilistic program. 
% via the raw random number calls of the simulator.
 But, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used in a probabilistic programming framework.
In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology modeling, multi-agent modeling and financial modeling. We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline how to do and provide examples of. We demonstrate our method on three models and provide n infected population simulator and demonstrate the hijacking on two important malaria population-based simulators OpenMalaria and EMOD.

>>>>

Probabilistic programming provides a way to perform statistical inference over simulations of events in a programmatic way. Thus, simulators by design are ideal programs for probabilistic programming systems~(PPSs).
However, within existing PPSs one would have to re-implement the simulator via the PPS language specification, which is inefficient and often not feasible due to the complexity of such scientific and industrial simulators.
However, recent work
by \cite{baydin2018efficient} demonstrated a pathway to turn a particular
type of event-based simulator into a probabilistic program. 
% via the raw random number calls of the simulator.
 But, this still meant that a large class of critically important population-based simulators could not be turned into probabilistic programs and as such could not be used in a probabilistic programming framework.
In this work, we extend that framework to encompass population-based simulators, a very large class of simulators that are used extensively across epidemiology modeling, multi-agent modeling and financial modeling. We enable simulators written in 13 different languages to be easily converted to probabilistic programs, for which we algorithmically outline how to do and provide examples of. We demonstrate our method on three real-world simulators for epidemiology.

\todo[inline]{We need to have probabilistic programming in the title to make sure we attract the correct reviewers}
